Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2013-10-14 Direction_JINST_v2.tex
Direction_JINST_v1.tex has the same content as Direction_v7.tex but is in JINST format. The references have been adjusted to match the JINST journal style. We use JHEP.bst as recommended in the JINST author manual. Direction_JINST_v2.tex has an additional section with three additional plots containing results for 1.4 and 2.1 MeV. Direction_JINST_v3.tex is another iteration of Direction_JINST_v2.tex and Direction_JINST_v4.tex is the version which Andrey, Henry, Matt and Christoph propose for publication. Some more details can be found in README.md --Christoph

\title{Measuring Directionality in Double-Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors}

@@

\section{Liquid Scintillator Detectors}
Liquid scintillators are `cocktails' of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiationless transitions. The first
excited state can de-excite through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

@@

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in FIG.~\ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd(Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr(Q=3.350~MeV)\cite{zr1},
$^{100}$Mo(Q=3.034~MeV)\cite{mo1}, $^{82}$Se(Q=2.995~MeV)\cite{qdot},
$^{116}$Cd(Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te(Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe(Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn(Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. As an alternative to single-atom-doping, recently nanocrystals 
formed by candidate isotopes have been explored as a dopant\cite{qdot,qdot2}.

\section{Geant4 Simulation}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a
\GEANT\cite{geant4one,geant4two} simulation has been constructed. The
simulation uses \GEANT~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

The initial studies described here are made with single 5~MeV electrons. 
Future studies will work on increasing the complexity and lowering 
the electron energies as is needed for $0\nu\beta\beta$ studies. Three 
effects primarily contribute to the timing of the scintillator detector 
system. First, the simulated travel time of the initial 5~MeV electron is typically between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

\section{Detector Timing}
\label{detector_timing_sec}

We first discuss results for the default simulation settings described
in the previous section. FIG.~\ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron (PE) detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector, with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cherenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed compared to
scintillation light. There is, however, significant overlap of the two
arrival time distributions.

@@

FIG.~\ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17-inch KamLAND PMTs\cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34.0$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel photodetector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}.
FIG.~\ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots; quantum dots can also provide a mechanism for introducing an 
isotope for studying double-beta decay.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34.0~ns timing cut.  Compared to
the default case shown in FIG.~\ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are longer than in the KamLAND emission spectrum.

However, advances in the production of commercial quantum dot samples
could yield quantum dots which have similar, single peak emission
shapes at lower wavelengths. This case has been simulated using the
same spectral shape of the measured core-shell quantum dot emission
but shifted to lower wavelengths such that the emission peak is
centered at 384~nm. This peak emission value has been measured for
other types of QDs, however with a much more pronounced 
tail\cite{qdot2}. The resulting PE time distribution shows improved
separation of Cherenkov and scintillation light compared to the
default simulation. After the 34.0~ns cut on the TTS-smeared PE time
we obtain a Cherenkov/scintillation ratio of $R_{C/S}=0.86$ (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

@@

\begin{figure*}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.4]{graphs/hDir_34ns.pdf}}
        \subfigure{\includegraphics[scale=0.4]{graphs/hVtx_34ns.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(\frac{p_x}{|\vec{p}|}, \frac{p_y}{|\vec{p}|},
        \frac{p_z}{|\vec{p}|})$, for the simulation of 1000 electrons
        (5~MeV). In the simulation the electrons are produced along the
        x-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation.}

\end{center}
\end{figure*}

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV. The electrons are
simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\frac{\vec{p}}{|\vec{p}|}$ = (1,0,0). FIG.~\ref{fig:reco}
shows the vertex reconstruction. The vertex is reasonably well
reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in FIG.~\ref{fig:reco}. It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms
used.

\section{Conclusions}
The ability to reconstruct direction in kiloton-scale scintillation
detectors would be a major technological advance for neutrino
experiments, especially those also
searching for neutrino-less double-beta decay. More generally, this technique could
be applied wherever scintillation-based detectors are used. A \GEANT~simulation of a simple spherical detector corresponding to a kiloton
of scintillator shows that timing on the order of 0.1~ns is required
to separate the directional Cherenkov light from the more abundant
scintillation light. The separation can be improved using
photodetectors with more red sensitivity and liquid scintillators with
a more narrow emission spectrum shifted to shorter
wavelengths. Furthermore, simple reconstruction algorithms adapted
from those for water Cherenkov detectors are able to reliably
reconstruct the position and direction of 5~MeV electrons. More
detailed simulation and advanced reconstruction algorithms will need
to be developed to move to lower energies and more complicated event
topography, such as those in neutrino-less double-beta decay, but the technique already
appears promising.

>>>>

\title{Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors}

@@

\section{Liquid scintillator detectors}
Liquid scintillators are `cocktails' of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiationless transitions. The first
excited state can de-excite through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

@@

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in figure \ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. As an alternative to single-atom-doping, recently nanocrystals 
formed by candidate isotopes have been explored as a dopant\cite{qdot,qdot2}.

\section{Geant4 simulation}
\label{sim_section}
In order to study the effects relevant to directional reconstruction
in liquid scintillators, a \GEANT \newline 
\cite{geant4one,geant4two} simulation
has been constructed. The simulation uses \GEANT~version 4.9.6 with the default liquid scintillator
optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene
(1,2,4-trimethylbenzene) and 1.52~g/l PPO (2,5-diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}. 
Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

@@

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure}

The studies in sections \ref{sim_section} to \ref{reconstruction_sec} are made with single 5~MeV electrons, lower energies are
discussed in section \ref{edep_section}. Future studies will work on increasing the complexity
of the event as is needed for $0\nu\beta\beta$ studies. Three
effects primarily contribute to the timing of the scintillator detector
system. First, the simulated travel time of the initial 5~MeV electron is typically between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

@@

\section{Detector timing}
\label{detector_timing_sec}

We first discuss results for the default simulation settings described
in the previous section. Figure \ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron (PE) detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector, with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cherenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed compared to
scintillation light. There is, however, significant overlap of the two
arrival time distributions.

@@

Figure \ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17-inch KamLAND PMTs\cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34.0$~ns (see
figure \ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel photodetector types.

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}.
Figure \ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

\section{Scintillator emission spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments\cite{qdot,qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots; quantum dots can also provide a mechanism for introducing an 
isotope for studying double-beta decay.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34.0~ns timing cut.  Compared to
the default case shown in figure \ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are longer than in the KamLAND emission spectrum.

However, advances in the production of commercial quantum dot samples
could yield quantum dots which have similar, single peak emission
shapes at lower wavelengths. This case has been simulated using the
same spectral shape of the measured core-shell quantum dot emission
but shifted to lower wavelengths such that the emission peak is
centered at 384~nm. This peak emission value has been measured for
other types of QDs, however with a much more pronounced 
tail\cite{qdot2}. The resulting PE time distribution shows improved
separation of Cherenkov and scintillation light compared to the
default simulation. After the 34.0~ns cut on the TTS-smeared PE time
we obtain a Cherenkov/scintillation ratio of $R_{C/S}=0.86$ (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.
\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_34ns.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_34ns.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (5~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation.}
\end{center}
\end{figure}

@@

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV, lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco}
shows the vertex reconstruction. The vertex is reasonably well
reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco}. It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms
used.

\section{Energy dependence}
\label{edep_section}
In the previous sections we presented results on single 5~MeV electrons.
In this section we study two lower energies, 1.4~MeV and 2.1~MeV. These
energies correspond to Q/2 for the double beta decay of
$^{116}$Cd and $^{48}$Ca, respectively\cite{cd1,biller}. The isotope
$^{116}$Cd was chosen because of its potential use in quantum-dot-doped
scintillators\cite{qdot,qdot2} and $^{48}$Ca was chosen
to cover the Q-value range of $0\nu\beta\beta$ candidate isotopes.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.5]{graphs/PEandR_vs_energy.pdf}
        \caption[]{The energy dependence of the mean number of PEs after the 34.0~ns time cut is shown for Cherenkov-induced PEs (black circles) and scintillation-induced PEs (black squares). The ratio between the mean number of Cherenkov-induced and scintillation-induced PEs is shown as blue triangles, values are given on the right y-axis. The statistical errors are too small to be seen. \label{Edep_NPE}}
        \end{center}
\end{figure}

The two additional simulation sets with 1.4 MeV and 2.1 MeV electrons were
generated using the default simulation configuration described in section
\ref{sim_section}. The PE time distribution for the default settings is shown
in figure \ref{time_plots_comparison} (a) for 5~MeV electrons. The shape
of the scintillation and Cherenkov spectra is similar for the lower energies (not
shown here). In figure \ref{Edep_NPE}, the energy-dependent mean number of
PEs per event after the 34.0~ns time cut is shown for Cherenkov-induced and
scintillation-induced PEs, as well as their ratio $R_{C/S}$. The mean number
of PEs from Cherenkov (scintillation) light is 21.8 (52.1), 38.4 (76.8)
and 108 (171) for 1.4~MeV, 2.1~MeV and 5~MeV, respectively. This gives the
ratios $R_{C/S}$ = 0.42, 0.50 and 0.63: The decrease in Cherenkov-induced
PEs is stronger than the decrease in scintillation-induced PEs as the energy
is lowered.

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hDir_pq_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs/hVtx_pq_1p4.pdf}}
        \caption[]{(Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons
        (1.4~MeV). In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. \label{low_energy_reco}}
\end{center}
\end{figure}
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/hCos_vs_E.pdf}
        \caption[]{Mean cosine of the angle between the true initial electron direction and the reconstructed direction, depending on the energy of the initial electrons. For each energy 1000 events have been simulated. Statistical errors are
shown. \label{Edep_angle}}
        \end{center}
\end{figure}

The reconstruction algorithms outlined in section \ref{reconstruction_sec} have also been applied
to the simulations at lower energies. Figure \ref{low_energy_reco} shows the results for 
the lowest simulated electron energy, 1.4~MeV. Most events are still reconstructed well, despite the lower
number of PEs and the decreased $R_{C/S}$. For 1.4~MeV electrons, the distribution RMS values for all three
reconstructed coordinates are smaller than 4.5~cm. The mean cosine of the angle between the true direction
and the reconstructed direction for different energies of the initial electron is shown in figure
\ref{Edep_angle}. The direction reconstruction performance is still promising for energies as low as 1.4~MeV.

\section{Conclusions}
The ability to reconstruct direction in kiloton-scale scintillation
detectors would be a major technological advance for neutrino
experiments, especially those also
searching for neutrino-less double-beta decay. More generally, this technique could
be applied wherever scintillation-based detectors are used. A \GEANT~simulation of a simple spherical detector corresponding to a kiloton
of scintillator shows that timing on the order of 0.1~ns is required
to separate the directional Cherenkov light from the more abundant
scintillation light. The separation can be improved using
photodetectors with more red sensitivity and liquid scintillators with
a more narrow emission spectrum shifted to shorter
wavelengths. Furthermore, simple reconstruction algorithms adapted
from those for water Cherenkov detectors are able to reliably
reconstruct the position and direction of electrons with energies in the few MeV range.
More detailed simulation and advanced reconstruction algorithms will need
to be developed to move to more complicated event
topography, such as those in neutrino-less double-beta decay, but the technique already
appears promising.
